### 2016-05-02 text/Chapter06.tex
Translational semantics for structural coeffects (Chapter 6)

\subsection{Structural indexed comonads}

@@

\begin{figure*}[!t]
The semantics is defined over a typing derivation:
%
\begin{equation*}
\hspace{-3em}
\begin{array}{ll}
\hspace{4em}\semdef
  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau}
  {\ident{counit}_{\cunit}}
& (\emph{var})
\\[1em]
\hspace{4.6em}\semdef
  {\coctx{()}{\alift{}} \vdash n : \ident{num} }
  {\ident{const}~n}
& (\emph{num})
\\[1.5em]
\hspace{1.40em}\semdeff
  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
  {f}
  {f \circ \ident{curry}~\ident{merge}_{\aclrd{\textbf{r}}, \cclrd{s}} }
& (\emph{abs})
\\[2em]
\hspace{-1.85em}\emdefff
  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
  {\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\begin{array}{l}\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\,\atimes\,(\cclrd{t} \aseq \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}\\[-0.3em]~\end{array}}
  {f}
  {g}
  {\hspace{-0.5em}\begin{array}{l}
  \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~g)\\[-0.3em]
  ~~\circ~~\ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}\\[-0.3em]
  \end{array}\hspace{-0.5em}}
& (\emph{app})
\\[1.5em]
\hspace{1.55em}\semdeff
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}
  {f}
  {f\circ\ident{snd}\circ \ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}}
& (\emph{weak})
\\[1.5em]
\hspace{0.5em}\mdeff
  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2\\[-0.25em]\qquad}
     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}{\aclrd{\textbf{r}}
          \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
  {f}
  {\begin{array}{l}\hspace{-0.5em}f\,\circ\,\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}},
     \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}\\[-0.25em]
     ~~(\ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}\circ\ident{swap}\circ\ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}})\hspace{-5em}~ \end{array} }
& (\emph{exch})
\\[1.5em]
\hspace{0em}\mdeff
  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}
      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}}
          \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \\[-0.25em] \qquad\vdash \subst{e}{z,y}{x} : \tau}\end{array}}
  {f}
  {\begin{array}{l}\hspace{-0.5em}f\,\circ\,\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}},
      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}\hspace{-0.5em} \\[-0.25em]~\end{array}}
& (\emph{contr})
\end{array}
\end{equation*}

\vspace{2em}
Assuming the following auxiliary operations:
%
\begin{equation*}
\begin{array}{rcl}
  \ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~f &=&
    \ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}} \;\circ\;
      \ident{id} \times(\ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} \,\circ\, f \times \ident{id} \,\circ\, \ident{split}_{\aclrd{\textbf{s}}, \aclrd{\textbf{t}}}) \;\circ\;
      \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}} \\[-0.3em]
  \ident{map}_\cclrd{r}}~f &=& \ident{cobind}_{\cunit, \cclrd{r}}~(f\circ\ident{counit}_\cunit) \\[-0.3em]
\end{array}
\end{equation*}
\begin{equation*}
\begin{array}{rcl}
  \ident{id}~x &=& x \\[-0.3em]
  \ident{const}~v &=& \lambda x.v \\[-0.3em]
  \ident{curry}~f~x~y &=& \lambda f.\lambda x.\lambda y.f~(x, y)\\[-0.3em]
  \ident{snd}~(x,y) &=& y\\[-0.3em]
  \ident{swap}~(x,y) &=& (y,x)\\[-0.3em]
  f \times g &=& \lambda (x, y).(f~x, g~y)\\[-0.3em]
  \ident{app}~(f, x) &=& f~x
\end{array}
\end{equation*}

\figcaption{Categorical semantics of the structural coeffect calculus}
\label{fig:struct-semantics}
\end{figure*}
% --------------------------------------------------------------------------------------------------

@@

\begin{theorem}[Correspondence]
In all of the typing rules of the structural coeffect system, the context annotations $\aclrd{\textbf{r}}$
and $\cclrd{s}$ of typing judgements $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ and function types
$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$  correspond to the indices of mappings $\ctyp{\aclrd{\textbf{r}}}{}$
and $\ctyp{\alift{\cclrd{s}}}{}$ in the corresponding semantic function defined
by $\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}$.
\end{theorem}
\begin{proof}
By analysis of the semantic rules in Figure~\ref{fig:struct-semantics}.
\end{proof}

@@

\section{Translation}
\label{sec:struct-transl}

@@

>>>>

\subsection{Structural indexed comonads}
\label{sec:struct-semantics-idxc}

@@

\begin{figure*}[!t]
The semantics is defined over a typing derivation:
%
\begin{equation*}
\hspace{-3em}
\begin{array}{ll}
\hspace{4em}\semdef
  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau}
  {\ident{counit}_{\cunit}}
& (\emph{var})
\\[1em]
\hspace{4.6em}\semdef
  {\coctx{()}{\alift{}} \vdash n : \ident{num} }
  {\ident{const}~n}
& (\emph{num})
\\[1.5em]
\hspace{1.40em}\semdeff
  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
  {f}
  {f \circ \ident{curry}~\ident{merge}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}} }
& (\emph{abs})
\\[2em]
\hspace{-1.85em}\emdefff
  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
  {\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\begin{array}{l}\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\,\atimes\,(\cclrd{t} \aseq \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}\\[-0.25em]~\end{array}}
  {f}
  {g}
  {\hspace{-0.5em}\begin{array}{l}
  \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~g)\\[-0.25em]
  ~~\circ~~\ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}\\[-0.25em]
  \end{array}\hspace{-0.5em}}
& (\emph{app})
\\[1.5em]
\hspace{1.55em}\semdeff
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}
  {f}
  {f\circ\ident{snd}\circ \ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}}
& (\emph{weak})
\\[1.5em]
\hspace{0.5em}\mdeff
  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2\\[-0.25em]\qquad}
     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}{\aclrd{\textbf{r}}
          \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
  {f}
  {\begin{array}{l}\hspace{-0.5em}f\,\circ\,\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}},
     \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}\\[-0.25em]
     ~~(\ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}\circ\ident{swap}\circ\ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}})\hspace{-5em}~ \end{array} }
& (\emph{exch})
\\[1.5em]
\hspace{0em}\mdeff
  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}
      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}}
          \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \\[-0.25em] \qquad\vdash \subst{e}{z,y}{x} : \tau}\end{array}}
  {f}
  {\begin{array}{l}\hspace{-0.5em}f\,\circ\,\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}},
      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~\ident{dup}_{\cclrd{s}, \cclrd{t}}\hspace{-0.5em} \\[-0.25em]~\end{array}}
& (\emph{contr})
\end{array}
\end{equation*}

\vspace{2em}
Assuming the following auxiliary operations:
%
\begin{equation*}
\begin{array}{rcl}
  \ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~f &=&
    \ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}} \;\circ\;
      \ident{id} \times(\ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} \,\circ\, f \times \ident{id} \,\circ\, \ident{split}_{\aclrd{\textbf{s}}, \aclrd{\textbf{t}}}) \;\circ\;
      \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}}
\end{array}
\end{equation*}
\begin{equation*}
\begin{array}{rcl}
  \ident{id}~x &=& x \\[-0.25em]
  \ident{const}~v &=& \lambda x.v \\[-0.25em]
  \ident{curry}~f~x~y &=& \lambda f.\lambda x.\lambda y.f~(x, y)\\[-0.25em]
  \ident{fst}~(x,y) &=& x\\[-0.25em]
  \ident{swap}~(x,y) &=& (y,x)\\[-0.25em]
  f \times g &=& \lambda (x, y).(f~x, g~y)\\[-0.25em]
  \ident{app}~(f, x) &=& f~x
\end{array}
\end{equation*}

\figcaption{Categorical semantics of the structural coeffect calculus}
\label{fig:struct-semantics}
\end{figure*}

@@

\begin{theorem}[Correspondence]
\label{thm:struct-correspondence}
In all of the typing rules of the structural coeffect system, the context annotations $\aclrd{\textbf{r}}$
and $\cclrd{s}$ of typing judgements $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ and function types
$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$  correspond to the indices of mappings $\ctyp{\aclrd{\textbf{r}}}{}$
and $\ctyp{\alift{\cclrd{s}}}{}$ in the corresponding semantic function defined
by $\sem{\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}$.
\end{theorem}
\begin{proof}
By analysis of the semantic rules in Figure~\ref{fig:struct-semantics}.
\end{proof}

@@

\section{Translational semantics}
\label{sec:struct-transl}

In the previous section, we used category theory to give a unified model capable of capturing
the semantics of our three context-aware language based on the structural coeffect calculus.
Although the categorical model is interesting on its own, we use it in the same way as in
Chapter~\ref{ch:semantics} -- to define a translation from source context-aware languages to a
simple target functional language. As for flat coeffect calculus, we show that the translation
produces well-typed programs in the target language. For a sample context-aware programming
language, we then show that well-typed programs (produced by the translation) do not get stuck.

This section mirrors the development presented in Chapter~\ref{ch:semantics} for flat coeffect
calculus. We extend the simple target language with additional constructs inspired by structural
indexed comonads (Section~\ref{sec:struct-transl-ext}), define the translation to the target
langauge (Section~\ref{sec:struct-transl-transl}) and prove safety of one sample language
(Section~\ref{sec:struct-transl-safety}) -- we choose structural dataflow to allow easy
comparison with the flat system.

\begin{figure}[t]
\paragraph{Language syntax.}
Given a structural coeffect algebra, extend the programming language syntax with the following
constructs:

\begin{equation*}
\begin{array}{rcl}
  e &  =  &\ldots \sep\ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~e_1~e_2 \sep \ident{counit}_{\cunit}~e \sep \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e
         \sep \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e \sep \ident{dup}_{\cclrd{r},\cclrd{s}}~e \\
  \tau &  = &\ldots \sep \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} \\
  K  & = &\ldots \sep \ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~\_~e \sep \ident{cobind}_{\cclrd{s},\aclrd{\textbf{r}}}~v~\_ \sep \ident{counit}_{\cunit}~\_ \\
    && \hspace{1.4em} \sep \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~\_
           \sep \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~\_ \sep \ident{dup}_{\cclrd{r},\cclrd{s}}~\_ \\
\end{array}
\end{equation*}

~

~

\paragraph{Typing rules.}
Given a structural coeffect algebra, add the typing rules:

\begin{equation*}
\begin{array}{l}
\tyrule{counit}
  {\Gamma \vdash e : \ctyp{\alift{\cunit}}{\tau}}
  {\Gamma \vdash \ident{counit}_{\cunit}~e : \tau} \\
\\
\tyrule{cobind}
  {\Gamma \vdash e_1 : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} \rightarrow \tau & \Gamma \vdash e_2 : \ctyp{\cclrd{s}\aseq\aclrd{\textbf{r}}}{\tau_1(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
  {\Gamma \vdash \ident{cobind}_{\cclrd{r}, \cclrd{s}}~e_1~e_2 : \ctyp{\alift{\cclrd{s}}}{\tau}} \\
  \\
\tyrule{merge}
  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \xtimes \ldots \xtimes \tau_k)} }
  {\Gamma \vdash  \ident{merge}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} }\\
\\
\tyrule{split}
  {\Gamma \vdash  e : \ctyp{\aclrd{\textbf{r}}\atimes\aclrd{\textbf{s}}}{(\tau_1 \xtimes \ldots \xtimes \tau_k)} }
  {\Gamma \vdash  \ident{split}_{\aclrd{\textbf{r}},\aclrd{\textbf{s}}}~e : \ctyp{\aclrd{\textbf{r}}}{(\tau_1 \xtimes \ldots \xtimes \tau_l)} \times \ctyp{\aclrd{\textbf{s}}}{(\tau_{l+1} \xtimes \ldots \xtimes \tau_k)}} \\
\\
\tyrule{dup}
  {\Gamma \vdash  e : \ctyp{\alift{\cclrd{r}\cpar\cclrd{s}}}{\tau} }
  {\Gamma \vdash  \ident{dup}_{\cclrd{r},\cclrd{s}}~e : \ctyp{\alift{\cclrd{r},\cclrd{s}}}{(\tau \xtimes \tau)} } \\
\end{array}
\end{equation*}

\caption{Comonadically-inspired extensions for structural coeffects}
\label{fig:struct-semantics-ext}
\end{figure}

\subsection{Comonadically-inspired language extensions}
\label{sec:struct-transl-ext}

In Section~\ref{sec:semantics-translation-target}, we defined the syntax, typing rules and
operational semantics for a simple functional programming language. We then extended it with
uninterpreted constructs inspired by the \emph{flat indexed comonad} structure and used it as the
translation target for the flat coeffect calculus. In this section, we take the same core language
and extend it with constructs inspired by the \emph{structural indexed comonad}.

Given a coeffect language with a structural coeffect algebra formed by
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq, \atimes$, we
extend the core functional language with operations shown in Figure~\ref{fig:struct-semantics-ext}.
The syntax extensions add comonadically-inspired operations that mirror those defined in
Section~\ref{sec:struct-semantics-idxc}. The typing for the operations corresponds to their
categorical counterparts.

We also include an uninterpreted type $\ctyp{\aclrd{\textbf{r}}}{\tau_1\xtimes\ldots\xtimes\tau_k}$,
which models a contextual (comonadic) value indexed by a vector of annotations. As in the
categorical model for structural coeffects, context consisting of multiple variables is not modelled
as ordinary tuple -- it can only be manipulated by the comonadically-inspired operations.
In the target language, this is done by defining the $\ctyp{\aclrd{\textbf{r}}}{}$ type over
zero or more underlying types. Here, our syntactic treatment differs slightly from the categorical
model where objects created by $\xtimes$ were first-class values. In Figure~\ref{fig:struct-semantics-ext},
we need to explicitly specify the types of individual components in typing rules for
(\emph{merge}), (\emph{split}) and (\emph{cobind}).

As with flat coeffects, the extensions described here are common for all concrete instances of
structural context-aware languages. For each concrete language, we need to provide values of type
$\ctyp{\aclrd{\textbf{r}}}{\tau}$ and reduction rules for comonadically-inspired operations.

\begin{figure*}[!t]
The translation is defined over a typing derivation:
%
\begin{equation*}
\begin{array}{ll}
\hspace{3.25em}\semdef
  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau}
  {\lambda\ctx.\ident{counit}_{\cunit}~\ctx}
& (\emph{var})
\\[1em]
\hspace{4em}\semdef
  {\coctx{()}{\alift{}} \vdash n : \ident{num} }
  {\lambda\ctx.n}
& (\emph{num})
\\[1.5em]
\hspace{0.8em}\semdeff
  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\atimes\,\alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
  {f}
  {\lambda\ctx.\lambda v.f~(\ident{merge}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}}}~(\ctx, v)) }
& (\emph{abs})
\\[2em]
\hspace{-2.5em}\emdefff
  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
  {\coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\begin{array}{l}\sem{\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\,\atimes\,(\cclrd{t} \aseq \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}\\[-0.25em]~\\[-0.25em]~\end{array}}
  {f}
  {g}
  {\hspace{-0.5em}\begin{array}{l}
  \lambda\ctx.~~\\[-0.25em]
    \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\aclrd{\textbf{r}}, \cclrd{t} \,\aseq\, \aclrd{\textbf{s}}}~\ctx\\[-0.25em]
    \quad f~\ctx_1~(\ident{cobind}_{\cclrd{t}, \aclrd{\textbf{s}}}~g~\ctx_2)
  \end{array} }
& (\emph{app})
\\[1.5em]
\hspace{0em}\emdeff
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\begin{array}{l}\sem{\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \alift{ \czero }} \vdash e : \tau}\\[-0.25em]~\\[-0.25em]~\\\end{array}}
  {f}
  {\hspace{-0.5em}\begin{array}{l}
  \lambda\ctx.~~\\[-0.25em]
    \quad \kvd{let}~(\ctx_1, \_) = \ident{split}_{\aclrd{\textbf{r}}, \alift{\czero}}~\ctx\\[-0.25em]
    \quad f~\ctx_1
  \end{array} }
& (\emph{weak})
\\[1.5em]
\hspace{0em}\mdeff
  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2\\[-0.25em]\qquad}
     {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}{\aclrd{\textbf{r}}
          \atimes \alift{\cclrd{t},\cclrd{s}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\\[-0.25em]~\end{array}}
  {\begin{array}{l}~\\[-0.25em]\hspace{-0.5em}f\end{array}}
  {\begin{array}{l}\hspace{-0.5em}
    \lambda\ctx.~f (\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{t},\cclrd{s}},\alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}~(\lambda\ctx'.\\[-0.25em]
     \quad \kvd{let}~(\ctx_1, \ctx_2) = \ident{split}_{\alift{\cclrd{t}}, \alift{\cclrd{s}}} \\[-0.25em]
     \quad \ident{merge}_{\alift{\cclrd{s}}, \alift{\cclrd{t}}}~(\ctx_2, \ctx_1)~))\end{array} }
& (\emph{exch})
\\[1.5em]
\hspace{-0.5em}\mdeff
  {\begin{array}{l}\sem{\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2\\[-0.25em]\qquad}
      {\aclrd{\textbf{r}} \atimes \alift{\cclrd{s},\cclrd{t}} \atimes \aclrd{\textbf{q}}} \vdash e : \tau}\end{array}}
  {\begin{array}{l}\sem{\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}}
          \atimes \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \atimes \aclrd{\textbf{q}}} \\[-0.25em] \qquad\vdash \subst{e}{z,y}{x} : \tau}\end{array}}
  {\begin{array}{l}~\\[-0.25em]\hspace{-0.5em}f\end{array}}
  {\begin{array}{l}\hspace{-0.5em}\lambda\ctx.f~(\ident{nest}_{\aclrd{\textbf{r}}, \alift{\cclrd{s}\,\cpar\,\cclrd{t}},
      \alift{\cclrd{s},\cclrd{t}}, \aclrd{\textbf{q}}}\\[-0.25em]
      \qquad \ident{dup}_{\cclrd{s}, \cclrd{t}}~\ctx)\hspace{-0.5em}\end{array}}
& (\emph{contr})
\end{array}
\end{equation*}

\vspace{2em}
Assuming the following auxiliary definition:
%
\begin{equation*}
\begin{array}{l}
  \ident{nest}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{s'}}, \aclrd{\textbf{t}}} = \lambda f.\lambda \ctx.\\[-0.25em]
  \qquad \kvd{let}~(\ctx_1, \ctx') = \ident{split}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s}}\atimes\aclrd{\textbf{t}}}~\ctx\\[-0.25em]
  \qquad \kvd{let}~(\ctx_2, \ctx_3) = \ident{split}_{\aclrd{\textbf{s}},\aclrd{\textbf{t}}}~\ctx'\\[-0.25em]
  \qquad \ident{merge}_{\aclrd{\textbf{r}}, \aclrd{\textbf{s'}}\atimes\aclrd{\textbf{t}}}~(\ctx_1, \ident{merge}_{\aclrd{\textbf{s'}}, \aclrd{\textbf{t}}}~(f~\ctx_2, \ctx_3))
\end{array}
\end{equation*}

\figcaption{Trnaslation from a structural coeffect calculus}
\label{fig:struct-translation}
\end{figure*}

\subsection{Comonadically-inspired translation}
\label{sec:struct-transl-transl}

When translating context-aware programs to the functional language, variable contexts become
values of comonadically-inspired data types containing a value for each variable in the context.
Function inputs become comonadically-inspired values containing exactly one variable. More formally:
%
\begin{equation*}
\begin{array}{rcl}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}} &=&
  \ctyp{\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}}{(\sem{\tau_1} \xtimes \ldots \xtimes \sem{\tau_n})} \\[0.5em]
\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} & = & \ctyp{\alift{\cclrd{r}}}{\sem{\tau_1}} \rightarrow \sem{\tau_2} \\[-0.2em]
\sem{\ident{num}} & = & \ident{num} \\
\end{array}
\end{equation*}

\noindent
The definition differs from the one for flat coeffects (Section~\ref{sec:semantics-translation-transl})
in that the comonadically-inspired data type takes multiple type parameters (separated using $\xtimes$),
rather than wrapping a regular tuple in the target language.

The translation rules are defined in Figure~\ref{fig:struct-translation}. As in the case of the flat
coeffect calculus, the definition directly follows the categorical semantics shown in
Figure~\ref{fig:struct-translation}. We expand the definitions so that the result is a valid
program in the target language rather than a composition of morphisms.

As with flat coeffects, the correspondence property of the semantics (Theorem~\ref{thm:struct-correspondence})
can now be adapted into well-typedness of the translation. Given a well-typed program in the
structural coeffect calculus, the translation produces a well-typed program in the target
language. This is true for all context-aware languages based on the structural coeffect calculus
and it provides us with the first part of type safety theorem. The second part is type safety
of the target language with concrete domain-specific extensions as discussed in
Section~\ref{sec:struct-transl-safety}.

\begin{theorem}[Well-typedness of the translation]
\label{thm:semantics-welltyped}
Given a typing derivation for a well-typed closed expression $\coctx{}{\alift{}} \vdash e : \tau$
written in a structural context-aware programming langugae that is translated to the target language as
(we write $\ldots$ for the omitted part of the translation tree):
%
\begin{equation*}
  \semdeff
    {~~(\ldots)~~}
    {\coctx{}{\alift{}} \vdash e : \tau }
    {(\ldots)}
    { ~~f }
\end{equation*}
%
Then $f$ is well-typed, \ie~in the target language: $\vdash f : \sem{\coctx{()}{\alift{}}} \rightarrow \sem{\tau}$.
\end{theorem}
\begin{proof}
By rule induction over the derivation of the translation. Given a judgement
$\coctx{x_1\!:\!\tau_1 \ldots x_n\!:\!\tau_n}{\aclrd{\textbf{c}}} \vdash e : \tau$ where
$\aclrd{\textbf{c}} = \alift{\cclrd{c_1},\ldots,\cclrd{c_n}}$, the translation
constructs a function of type $\ctyp{\aclrd{\textbf{c}}}(\sem{\tau_1}\xtimes\ldots\xtimes\sem{\tau_n}}) \rightarrow \sem{\tau}$.

\vspace{0.5em}\noindent\hangindent=0.6cm
Case (\emph{var}): $\aclrd{\textbf{c}}=\alift{\cunit}$ and so ${\ident{counit}_\cunit}~\ctx$ is well-typed.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{num}): $\tau = \ident{num}$ and so the body $n$ is well-typed.

%
%
%
%
TODO
%
%
%
%
%

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{abs}): The type of $\ctx$ is $\ctyp{\cclrd{r}}(\ldots)$ and the type of $v$ is $\ctyp{\cclrd{s}}\tau_1$,
  calling $\ident{merge}_{\cclrd{r},\cclrd{s}}$ and reassociating produces
  $\ctyp{\cclrd{r}\czip\cclrd{s}}(\ldots)$ as expected by $f$.

\vspace{0.25em}\noindent\hangindent=0.6cm
Case (\emph{app}): After applying $\ident{split}_{\cclrd{r}, \cclrd{s}\cseq\cclrd{t}}$, the types of
  $\ctx_1, \ctx_2$ are $\ctyp{\cclrd{r}}(\ldots)$ and $\ctyp{\cclrd{s}\cseq\cclrd{t}}(\ldots)$, respectively.
  $g$ requires $\ctyp{\cclrd{s}}(\ldots)$ and so the result of $\ident{cobind}_{\cclrd{s},\cclrd{t}}$ is
  $\ctyp{\cclrd{t}}\tau_1$ as required by $f$.
\end{proof}

@@

\newpage
